Aircraft Wildlife Collisions

### 2021-07-21 draft_final.Rmd
Formatting of variables in text and some plot labelling.

We then finish with a summary of our findings and further research questions that could possibly be further analyzed in the future.

@@

birds_struck: We are using the "remarks" variable to check for spare parts/hints in text from which we can determine that no wild life has been struck. By doing this we were able to decrease the missing values by 19 (from 39 to 20).

birds_seen: We assume that missing values in the "birds_seen" variable indicate that the pilot did not see wildlife before the strike occurred. Thus, we set all *NA* values in the "birds_seen" variable as "None". This leads to `r sum(is.na(birds$birds_seen))` values which are set as "None".

@@

The dataset contains `r length(unique(birds$atype))` different aircraft models. We extracted the overall maximum speed and height for each aircraft to get an overview of the aircraft models and the relationship between speed and height. The following left plot shows the Top 50 aircraft models that are causing the most strikes and on the right you can see the speed height relationship of the Top 50 aircrafts colored by their ac_mass.

@@

The left plot shows that especially the **B-737 class** (B-737-200,B-737-300) causes the most strikes together with MD-80 and B-727. "B" stands here for Boeing and MD-80 is an aircraft from McDonnell Douglas. The Top model causing strikes is the B-737-300 which caused in total `r top50[1,"n"]` strikes. The B-737 is a narrow-body airliner produced by Boeing at its Renton Factory in Washington and was developed for short and thin routes. Unfortunately we do not have the amount of flights and duration of the flight in the data, so we can not do further analysis on this. But we assume, that the aircraft models with higher amount of strikes had also more flights and or longer flights.

From the right plot we can see, that aircraft **models with higher max speed are also able to fly higher**. This can be also seen from the linear model, that has been estimated by geom_smooth. The correlation of max_speed and height for the Top50 models is: `r round(cor(top50$max_speed, top50$max_height),3)`, which is positive and quite high. In addition we can see via the coloring by ac_mass that **lighter aircrafts tend to have less max_speed and max_height**. We can also see that the lighter aircrafts under 5700 kg or even 27000 kg are less prominent in the Top50. The group with an **aircraft mass between 27001-272000 kg is the most prominent group** under the Top50 aircrafts. Two aircraft models (B-747-1/200 and B-747-400) with ac_mass of more than 272000 kg are also Boeings and present in the Top50 which are the only aircrafts of that weight class overall.

The next plot shows again the max_speed by max_height splitted by number of engines and colored by ac_mass but for all `r nrow(machines)` aircraft models. Since some models had very few strikes it can happen that they have been assigned an max_height of about 0.

@@

From the plot we can see, that again higher max_speed causes higher max_height and most of the aircraft models have two engines. We can also see that lighter machines tend to have less engines: 
* first group includes nearly only machines with aircraft mass = 1 (less than 2250 kg), 
* second group is mixed with aircraft mass 2-4 (2251-272000 kg), 
* the last two groups include mostly aircrafts with mass = 5 (27001-272000 kg).

@@

```{r echo=FALSE, fig.asp=0.5, fig.align="center"}
# create some fake data (2x2, since we're building a confusion matrix) 
dat <- matrix(table(machines$num_engs, machines$ac_mass), nrow = 4, dimnames=list(c(1:4), c(1:5)))
# now set the names *of the dimensions* (not the row/colnames)
names(dimnames(dat)) <- c("num_engs", "ac_mass")
dat
```
Interesting is that for aircrafts with one engine, there is no model with a speed greater 200 knots and height greater 10000 foot. So we could conclude from that this is the maximum height and speed possible for aircrafts with one engine.

@@

# 3. Analysis of influences for birds_struck, birds_seen and effect
In the following analysis we will go more into detail about the different conditions regarding the amount of birds that have been struck (variable: birds_struck), the amount of birds that have been seen by the pilot before a strike happened (variable: birds_seen), and the different effects a strike had on the aircraft (variable: effect). Most of the time we will use barplots with position fill, mosaicplot and spineplots as we want to compare proportions and check for dependencies. We will first start with the analysis of strikes.

@@

The following plot on the left shows the amount of birds struck out birds_seen. On the top right we can see the amount of collisions by each phase of flight and on the bottom right we can see the amount of collisions by time of day. For coloring we use the birds_struck variable.

```{r echo=FALSE, fig.asp=0.5}
birds_strike_phase <- ggplot(birds[!is.na(birds$phase_of_flt) & !is.na(birds$birds_struck) & as.integer(birds$phase_of_flt)>2,]) +
  aes(x = phase_of_flt, fill=birds_struck) +
  scale_fill_brewer(palette="Blues") +
  geom_bar() +
  coord_flip() +
  theme(legend.position="none", 
        axis.title.y=element_blank(),
        axis.title.x=element_blank(),
        plot.subtitle = element_text(hjust =1)) +
  theme.title+
  labs(xlab = "phase of flight",
       subtitle = "Number of collisions by phase of flight")

# Most collisions during the day #genauer betrachten
birds_strike_time <- ggplot(birds[!is.na(birds$time_of_day) & !is.na(birds$birds_struck),]) + 
  aes(x=time_of_day, fill=birds_struck) + 
  scale_fill_brewer(palette="Blues") +
  geom_bar() +
  coord_flip() +
  theme(legend.position="none",
        axis.title.y=element_blank(),
        axis.title.x=element_blank(),
        plot.subtitle = element_text(hjust =1)) +
  theme.title+
  labs(xlab = "",
       subtitle = "Number of collisions by time of day     ")

@@

The mosaicplot shows that the largest group of birds seen is "None" followed by "2-10" and "11-100". So **most often we do not see birds before a strike** happens and most often 1 bird is struck. But if we see something, than according to the data we only see birds in groups, where larger groups of birds are less likely than smaller ones. If the pilot sees birds, **quite often less birds were hit than seen**. For example if the pilot sees 2-10 birds, he hits about 40% of the time less birds and for the group birds seen = 11-100, about 85% of the time the pilot hits less birds than seen. The event of hitting less birds than seen will later be analyzed more in detail in chapter 3.3.

@@

From the bottom right plot we can see that strikes happen most often during the Day with over 10500 collisions followed by Night with about 4500. Interesting to see is that strikes with larger groups of birds (11-100) happpen mostly during the Day (small values are not visible in the plot).

We will now have a look on the influence of the sky variable for the different amounts of birds struck.

@@

From the plot on the left we can see that with **increasing number of birds_struck the proportion of overcast increases** and the proportion of no clouds decreases. The proportion of some clouds is more or less the same across all groups. From that we can conclude that the higher the amount of birds struck the more clouds and thus we would assume that birds are flying in bigger groups more often when it is cloudy or the pilot as well as the birds are not able to see each other because of the clouds and so nobody can try to prevent the collision by dodging the other one.

On the right side we can see a subset of the birds_struck groups, which focuses on the minor categories in our dataset. From that we can see that the aircraft mass has an influence on the amount of birds struck. For the lightest aircraft class up to 2250 kg the proportion of 0 birds struck is very high, which means that the pilot has seen a bird and was likely to hit one, but he did not. For this category we would expect that lighter machines are able to initiate evasive maneuvers as they are more agile. The relationship of striking less birds than seen will be analyzed more in detail in chapter 3.3. 
It also stands out, that the proportion of striking over 100 birds is significantly higher for the heaviest aircrafts above 272000 kg. This will most likely be due to the fact that heavier machines are also larger in size and thus they will hit more birds when flying through a flock of birds compared to lighter and thus smaller machines.

@@

We will now have a look on the height variable together with phase of flight. The top plot shows a boxplot of height for flight phases, which are above ground. We defined Climb, En Route, Descent, Approach as phases which are above ground level. On the bottom plot we will then have a closer look on the height for the flight phases "En Route" and "Descent" by visualizing the density.

```{r echo=FALSE, fig.asp=0.5, warning=FALSE}
above_ground <- birds[as.integer(birds$phase_of_flt)>3 & as.integer(birds$phase_of_flt)<8 & !is.na(birds$phase_of_flt),]
height_by_phase <- ggplot(above_ground) +
  aes(x = phase_of_flt, y = height, fill=phase_of_flt) +
  stat_boxplot(geom="errorbar", width=0.4) +
  scale_fill_brewer(palette="Pastel2") +
  geom_boxplot(outlier.size = .9) +
  coord_flip() +
  theme(axis.title.y = element_blank())+
  theme.title+
  labs(title = "Height distribution for flight phases above ground")+
  ylim(0,20000)
# Heights during Descent are on average higher than 
# hypothesis: Collisions happen more often during Descent and are therefore more prominent in the data set?

#plot height density for Descent and EnRoute
density <- ggplot(birds[birds$phase_of_flt== "Descent" | birds$phase_of_flt=="En Route",]) +
  aes(x = height, group=phase_of_flt,fill=phase_of_flt ) +
  scale_fill_manual(values=brewer.pal(4,"Pastel2")[2:3]) +
  geom_density(adjust=1.5,alpha=.8) +
  xlim(0,20000)

@@

The height distribution for Climb and Approach are very similar and their medians are very close to the ground. 
The other two boxplots for "En Route" and "Descent" are more far away from ground and show more variance in the data. Thus, we will focus these two. Interesting here is that the 25% quantile of the "Descent" Boxplot is above the one from "En Route", same for the median and 75% quantile. We will investigate more the differences of the two distributions by using a density plot.

Both flight phases have the highest density for lower heights up to about 5000 foot. The curve for En Route is much steeper, but a little more narrow compared to "Descent". Both density cross at about 3750 which is the highest point of "Descent" phase. From that height upwards the probability to strike a birds is higher when being in Descent phase. This will be most likely because of the angle of the aircraft at which it is flying. In phase En Route it will be most likely parallel to the ground, so the height of the airplane does not change during that phase, but in "Descent" phase the aircraft has an negative angle and is loosing height to prepare for example for Approach phase.

We are performing a t.test to compare the two group height ("Descent" and "En Route"), so that we can confirm that both distributions are unequal.

@@

The T-Test also shows that the group means are different from each other and that the **mean in group "Descent" is higher** than mean of "En Route". One would actually expect that the height is smaller than in "En Route" because the Descent phase happens after the "En Route" phase, where the aircraft is loosing height to prepare for approach or during flight in general. So since all aircrafts have to be in the phase "En Route" before going over to "Descent" we can say that it is more likely to hit a bird in greater height when being in phase "Descent" rather than in "En Route". The main difference between these flight phases is that the aircraft is not flying parallel to the ground when being in phase "Descent". So the **angle of the aircraft during the flight is most likely having an influence on hitting birds**. And especially a negative angle increases the probability of hitting birds, based on the findings here and from chapter 3.1.1 by comparing the number of collisions by phase of flight.

@@

We will now have another look on the density of height, but this time split by the different numbers of birds involved in the collision. The following plot shows exactly this.

```{r echo=FALSE, warning=FALSE, fig.asp=0.5}
#could include only birds from vys categorization
#plot height density for Descent and EnRoute
density <- ggplot(above_ground[!is.na(above_ground$birds_struck),]) +
  aes(x = height, group=birds_struck,fill=birds_struck)+
  geom_density(adjust=1.5,alpha=.6) +
  scale_fill_brewer(palette="Blues") +
  theme(legend.position = "right") + 
  theme.title+
  xlim(0, 3500) +
  labs(title = "Density of height for flight phases above ground", subtitle = "for height less than 3500 foot, by birds_struck")
density

@@

The density plot shows that the probability for striking a bird is highes near ground level. This plot shows a **very steep and narrow peek at the beginning at about 100 foot** for all categories of birds_struck greater 0. So most of the collisions happen near ground level, which we have already seen at the beginning of chapter 3.1.1 that the flight phases Take-Off, Climb, Approach and Landing Roll have the highest number of strikes (see 3.1 Number of collisions by pflight phase). For birds_struck=0 the distribution is a little bit different from the others with a density of about 0.0016 from 200 foot to 500 foot. Remarkable is also the **peek for "Over 100" birds struck at a height of about 1800 foot**. For the "Over 100" category we have the lowest amount of observations in the dataset but nevertheless 4 out of 8 strikes happened here. So for this category we have the about the same probability of hitting birds near ground level and at about 1800 foot. One of the involved species here was a **flock of about 500 gulls**. Unfortunately the other observations had missing values in the variable "species", so we can not get more information about the species involved.

@@

There originally were only two levels in the data available for birds seen and we added one level for the missing values. The two plots at the top show the proportion of birds seen by the time of day (left plot) and the cloudiness of the sky (right plot). At the bottom you can see two spineplots that show the two continuous variables height and speed by the number of birds seen.

@@

birds_seen_plt + birds_seenNo_sky
```
```{r echo=FALSE, fig.asp=0.4, fig.width = 7, fig.show="hold", fig.align = 'left'}
par(mfrow=c(1,2))
spineplot(birds$height, birds$birds_seen, ylab="", yaxlabels = "", xlab = "height", col = rev(brewer.pal(n=3,name="BuGn")))
title(main = "Birds seen by height", cex.main=0.8, font.main= 3, col.main = "gray50")

spineplot(birds$speed, birds$birds_seen,ylab="",yaxlabels = "", xlab = "speed",
          col = rev(brewer.pal(n=3,name="BuGn")))
title(main = "Birds seen by speed", cex.main=0.8, font.main= 3, col.main = "gray50")
```

Time of Day: Based on the assumption that missing values in the data for birds_seen stands for the fact that the pilot did not see wildlife before a strike happened, the average proportion of birds seen before a strike happened is 1-(0.65+0.69+0.65+0.87)/4=0.285. So most often pilots do not see birds before a collision. The proportion of overall **birds seen at night is much lower** with just 0.13 compared to the other times (Dawn, Day, Dusk).

Sky: The proportion of **wildlife seen increases with cloudiness**. This is a **contradiction** to what we would have expected. We would have thought that with increasing cloudiness the amount of wildlife/birds seen decreases, due to the fact that the pilot can not see as good as when there are less or no clouds. So one reason for this could be that most of the pilots aren't within the area of clouds when the collision happens, but have recorded the cloudiness of the sky. Another reason could be that birds fly more often when its cloudy.

From the spineplots we can see again that it is less likely to see birds before a collision, but the higher the speed & height, its becoming even less likely to see birds before a collision. For height greater 2000 foot and for speed greater 140 knots, the probability to see birds before a collision decreases.

@@

In the following we will have a closer look on a self-generated variable, which compares the amount of birds seen with the amount of birds_struck. It will be true if the number of birds struck is smaller than birds seen and false if not. In the beginning (chapter 3.1.1) we have already seen, that the number of birds_seen does not always infer that the same amount of birds were struck. By using this variable, we want to have a look at what influences the proportion of striking less birds than seen.

The plot on the left shows the proportion of less strikes than seen by phase of flight. On the top right we can see a comparison of the proportion of less strikes by the aircraft mass and on the bottom right we compare it by different number of engines.

```{r echo=FALSE, fig.asp=0.3}
#Create new column
#calculate only when seen is not na
seen_vs_strike <- birds[birds$birds_seen!="None" & !is.na(birds$birds_struck),]
seen_vs_strike$birds_struck <- factor(seen_vs_strike$birds_struck)
seen_vs_strike$birds_seen <- factor(seen_vs_strike$birds_seen)
seen_vs_strike$less_strike_than_seen <- as.factor(as.integer(seen_vs_strike$birds_struck) < as.integer(seen_vs_strike$birds_seen)+2)

#TODO: Why is barplot different from ggplot???
#barplot(prop.table(table(seen_vs_strike$less_strike_than_seen,seen_vs_strike$effectYes),2), xlab = #"less_strike_than_seen")
#prop.table(table(seen_vs_strike$less_strike_than_seen,seen_vs_strike$effectYes),2)

less_strike_sums <- rep(addmargins(table(seen_vs_strike$less_strike_than_seen, seen_vs_strike$phase_of_flt),1)[3,3:8],each=2)

# phase of flight by birdsstruck < birds seen
seen_vs_strike_phase <- ggplot(seen_vs_strike[!is.na(seen_vs_strike$phase_of_flt) & as.integer(seen_vs_strike$phase_of_flt)>2,]) +
  aes(x=phase_of_flt, fill=less_strike_than_seen) +
  scale_fill_manual(values = color_pal_TF) +
  geom_bar(position = "fill") +
 # geom_text(aes(label=..count..),
 # geom_text(aes(label=ifelse(round(..count../less_strike_sums[2:16],2) >
  geom_text(aes(label=ifelse(round(..count../less_strike_sums,2) >
  #0.1,round(..count../less_strike_sums[2:16],2) , "")),
                               0.1,round(..count../less_strike_sums,2) , "")),
            stat='count',
            size = 3,
            position=position_fill(vjust = 0.5))+
  #coord_flip() +
  theme(legend.position="right", axis.title.y = element_blank())+
  theme.title+
  labs(title="Variables influencing less strikes than seen",
       subtitle = "Proportion of less strike than seen \nby phase of flight",
       fill='less strike \nthan seen'
       )

#ac_mass by less-strike-than-seen (MOSAIC)
seen_vs_strike_ac_mass <- ggplot(seen_vs_strike[!is.na(seen_vs_strike$ac_mass),]) +
  geom_mosaic(aes(x=product(ac_mass_renamed), fill=less_strike_than_seen)) +
  scale_fill_manual(values = color_pal_TF)+
  theme(legend.position="none", axis.text.y=element_blank(), axis.ticks.y = element_blank(), axis.text.x = element_text(angle = 90))+
  ylab("") + xlab("ac_mass in tons")

#num_engs by less-strike-than-seen (MOSAIC)
seen_vs_strike_num_engs <- ggplot(seen_vs_strike[!is.na(seen_vs_strike$num_engs),]) +
  geom_mosaic(aes(x=product(num_engs), fill=less_strike_than_seen)) +
  scale_fill_manual(values = color_pal_TF)+
  theme(legend.position="none", axis.text.y=element_blank(),axis.ticks.y = element_blank())+
  ylab("")

>>>>

We then finish with a summary of our findings and further research questions that could possibly be analyzed in the future.

@@

*birds_struck*: We are using the *remarks* variable to check for spare parts/hints in text from which we can determine that no wild life has been struck. By doing this we were able to decrease the missing values by 19 (from 39 to 20).

*birds_seen*: We assume that missing values in the *birds_seen* variable indicate that the pilot did not see wildlife before the strike occurred. Thus, we set all *NA* values in the *birds_seen* variable as *None*. This leads to `r sum(is.na(birds$birds_seen))` values which are set as *None*.

@@

The dataset contains `r length(unique(birds$atype))` different aircraft models. We extracted the overall maximum speed (*max_speed*) and height (*max_height*) for each aircraft to get an overview of the aircraft models and the relationship between speed and height. The following left plot shows the top 50 aircraft models that are causing the most strikes and on the right you can see the speed height relationship of the top 50 aircrafts colored by their *ac_mass*.

@@

The left plot shows that especially the **B-737 class** (including B-737-200 and B-737-300) causes the most strikes together with MD-80 and B-727. "B" stands here for Boeing and MD-80 is an aircraft from McDonnell Douglas. The model causing the most strikes is the B-737-300 which caused in total `r top50[1,"n"]` strikes. The B-737 is a narrow-body airliner produced by Boeing at its Renton Factory in Washington and was developed for short and thin routes. Unfortunately we do not have the amount of flights and duration of the flight in the data, so we can not do further analysis on this. But we assume, that the aircraft models with higher amount of strikes had also more flights and or longer flights.

From the right plot we can see, that aircraft **models with higher *max_speed* are also able to fly higher**. This can be also seen from the linear model, that has been estimated by geom_smooth. The correlation of *max_speed* and *max_height* for the top 50 models is: `r round(cor(top50$max_speed, top50$max_height),3)`, which is positive and quite high. In addition we can see via the coloring by *ac_mass* that **lighter aircrafts tend to have less max_speed and max_height**. We can also see that the lighter aircrafts under 5700 kg or even 27000 kg are less prominent in the top 50. The group with an **aircraft mass between 27001-272000 kg is the most prominent group** under the top 50 aircrafts. Two aircraft models (B-747-1/200 and B-747-400) with *ac_mass* of more than 272000 kg are also Boeings and present in the top 50 which are the only aircrafts of that weight class overall.

The next plot shows again the *max_speed* by *max_height* splitted by number of engines and colored by *ac_mass* but for all `r nrow(machines)` aircraft models. Since some models had very few strikes it can happen that they have been assigned an *max_height* of about 0.

@@

From the plot we can see, that again higher *max_speed* causes higher *max_height* and most of the aircraft models have two engines. Interesting is that for aircrafts with one engine, there is no model with a speed greater 200 knots and height greater 10000 foot. So we could conclude from that this is the maximum height and speed possible for aircrafts with one engine. We can also see that lighter machines tend to have less engines:

* first group includes nearly only machines with aircraft mass less than 2250 kg, 
* second group is mixed with aircraft mass between 2251-272000 kg, 
* the last two groups include mostly aircrafts with a mass of 27001-272000 kg.

@@

```{r echo=FALSE, fig.asp=0.5, fig.align="center"}
# create some fake data (2x2, since we're building a confusion matrix) 
dat <- matrix(table(machines$num_engs, machines$ac_mass), nrow = 4, dimnames=list(c(1:4), c(1:5)))
# now set the names *of the dimensions* (not the row/colnames)
names(dimnames(dat)) <- c("num_engs", "ac_mass")
dat
```

@@

# 3. Analysis of influences for birds_struck, birds_seen and effect
In the following analysis we will go more into detail about the different conditions regarding the amount of birds that have been struck (variable: *birds_struck*), the amount of birds that have been seen by the pilot before a strike happened (variable: *birds_seen*), and the different effects a strike had on the aircraft (variable: *effect*). Most of the time we will use barplots with position fill, mosaicplot and spineplots as we want to compare proportions and check for dependencies. We will first start with the analysis of strikes.

@@

The following plot on the left shows the amount of birds struck out of birds seen. On the top right we visualized the number of collisions by each phase of flight and on the bottom right the plot shows the amount of collisions by time of day. For coloring we use the *birds_struck* variable.

```{r echo=FALSE, fig.asp=0.5}
birds_strike_phase <- ggplot(birds[!is.na(birds$phase_of_flt) & !is.na(birds$birds_struck) & as.integer(birds$phase_of_flt)>2,]) +
  aes(x = phase_of_flt, fill=birds_struck) +
  scale_fill_brewer(palette="Blues") +
  geom_bar() +
  coord_flip() +
  theme.title+
  theme(legend.position="none",
        axis.title = element_text(vjust=-10),
        axis.title.y=element_blank(),
        axis.title.x=element_blank(),
        plot.subtitle = element_text(hjust =4)) +
  labs(xlab = "phase of flight",
       subtitle = "Number of collisions by phase of flight")

# Most collisions during the day #genauer betrachten
birds_strike_time <- ggplot(birds[!is.na(birds$time_of_day) & !is.na(birds$birds_struck),]) + 
  aes(x=time_of_day, fill=birds_struck) + 
  scale_fill_brewer(palette="Blues") +
  geom_bar() +
  coord_flip() +
  theme.title+
  theme(legend.position="none",
        axis.title.y=element_blank(),
        axis.title.x=element_blank(),
        plot.subtitle = element_text(hjust =3)) +
  labs(xlab = "",
       subtitle = "Number of collisions by time of day     ")

@@

The mosaicplot shows that the largest group of birds seen is "None" followed by "2-10" and "11-100". So **most often we do not see birds before a strike** happens and most often only 1 bird is struck. But if the pilot sees something, than according to the data he only sees birds in groups, where larger groups of birds are less likely than smaller ones. If the pilot sees birds, **quite often less birds were hit than seen**. For example if the pilot sees 2-10 birds, he hits about 40% of the time less birds and for the group *birds_seen* between 11-100, about 85% of the time the pilot hits less birds than seen. The event of hitting less birds than seen will later be analyzed more in detail in chapter 3.3.

@@

From the bottom right plot we can see that strikes happen most often during the day with over 10500 collisions followed by night with about 4500. Interesting to see is that strikes with larger groups of birds (11-100) happen mostly during the day (small values are not visible in the plot).

We will now have a look on the influence of the sky variable for the different amounts of *birds_struck* and analyze the influence of the aircraft mass.

@@

From the left plot we can see that with **increasing number of *birds_struck* the proportion of overcast increases** and the proportion of no clouds decreases. The proportion of some clouds is more or less the same across all groups. From that we can conclude that the higher the amount of birds struck the more clouds and thus we would assume that birds are flying in bigger groups more often when it is cloudy or the pilot as well as the birds are not able to see each other because of the clouds and so nobody can try to prevent the collision by dodging the other one.

On the right side we can see a subset of the *birds_struck* groups, which focuses on the less frequent categories in our dataset. From that we can see that the aircraft mass has an influence on the amount of birds struck. For the lightest aircraft class up to 2250 kg the proportion of 0 birds struck is very high, which means that the pilot has seen a bird and was likely to hit one, but he did not. For this category we would expect that lighter machines are better able to initiate unplanned evasive maneuvers as they are more agile. The relationship of striking less birds than seen will be analyzed more in detail in chapter 3.3. 
It also stands out, that the proportion of striking over 100 birds is significantly higher for the heaviest aircrafts above 272000 kg. This will most likely be due to the fact that heavier machines are also larger in size and thus they will hit more birds when flying through a flock of birds compared to lighter and thus smaller machines.

@@

We will now have a look on the height variable together with phase of flight. The top plot shows a boxplot of height for flight phases, which are above ground. We defined Climb, En Route, Descent and Approach as phases which are above ground level. On the bottom plot we will then have a closer look on the height for the flight phases "En Route" and "Descent" by visualizing the density.

```{r echo=FALSE, fig.asp=0.5, warning=FALSE}
above_ground <- birds[as.integer(birds$phase_of_flt)>3 & as.integer(birds$phase_of_flt)<8 & !is.na(birds$phase_of_flt),]
height_by_phase <- ggplot(above_ground) +
  aes(x = phase_of_flt, y = height, fill=phase_of_flt) +
  stat_boxplot(geom="errorbar", width=0.4) +
  scale_fill_brewer(palette="Pastel2") +
  geom_boxplot(outlier.size = .9) +
  coord_flip() +
  theme(axis.title.y = element_blank())+
  theme.title+
  labs(title = "Height distribution for flight phases above ground")+
  ylab("height in foot")+
  ylim(0,20000)
# Heights during Descent are on average higher than 
# hypothesis: Collisions happen more often during Descent and are therefore more prominent in the data set?

#plot height density for Descent and EnRoute
density <- ggplot(birds[birds$phase_of_flt== "Descent" | birds$phase_of_flt=="En Route",]) +
  aes(x = height, group=phase_of_flt,fill=phase_of_flt ) +
  scale_fill_manual(values=brewer.pal(4,"Pastel2")[2:3]) +
  geom_density(adjust=1.5,alpha=.8) +
  xlab("height in foot")+
  xlim(0,20000)

@@

The height distribution for "Climb" and "Approach" are very similar and their medians are very close to the ground. 
The other two boxplots for "En Route" and "Descent" are more far away from ground and show more variance in the data. Thus, we will focus on these two. Interesting here is that the 25% quantile of the "Descent" boxplot is above the one from "En Route", same for the median and 75% quantile. We will investigate more the differences of the two distributions by using a density plot.

Both flight phases have the highest density for lower heights up to about 5000 foot. The curve for "En Route" is much steeper, but a little more narrow compared to "Descent". Both density cross at about 3750 foot, which is the highest point of "Descent" phase. From that height upwards the probability to strike a birds is higher when being in "Descent" phase. This will be most likely because of the angle of the aircraft at which it is flying. In phase "En Route" it will be most likely parallel to the ground, so the height of the airplane does not change much during that phase, but in "Descent" phase the aircraft has an negative angle and is loosing height to prepare for example for "Approach" phase.

We are performing a t-test to compare the two group heights ("Descent" and "En Route"), so that we can confirm that both distributions are not equal.

@@

The t-test also shows that the **group means are different** from each other, because the p-value is less than 5% or even 1% level. The **mean in group "Descent" is higher** than mean of "En Route". One would actually expect that the height is smaller than in "En Route" because the Descent phase happens after the "En Route" phase, where the aircraft is loosing height to prepare for approach or during flight in general. So since all aircrafts mostly have to be in the phase "En Route" before going over to "Descent" we can say that it is more likely to hit a bird in greater height (greater 3750 foot) when being in phase "Descent" rather than in "En Route". The main difference between these flight phases is that the aircraft is not flying parallel to the ground when being in phase "Descent". So the **angle of the aircraft during the flight is most likely having an influence on hitting birds**. And especially a negative angle increases the probability of hitting birds, based on the findings here and from chapter 3.1.1 where we found out, that the "Approach" phase has the highest amount of strikes.

@@

We will now have another look on the density of *height*, but this time split by the different numbers of birds involved in the collision. The following plot shows exactly this.

```{r echo=FALSE, warning=FALSE, fig.asp=0.5}
#could include only birds from vys categorization
#plot height density for Descent and EnRoute
density <- ggplot(above_ground[!is.na(above_ground$birds_struck),]) +
  aes(x = height, group=birds_struck,fill=birds_struck)+
  geom_density(adjust=1.5,alpha=.6) +
  scale_fill_brewer(palette="Blues") +
  theme(legend.position = "right") + 
  theme.title+
  xlim(0, 3500) +
  labs(title = "Density of height for flight phases above ground", subtitle = "for height less than 3500 foot, by birds_struck")+
  xlab("height in foot")
density

@@

The density plot shows that the probability for striking a bird is highest near ground level. The density starts with a **very steep and narrow peek at the beginning at about 100 foot** for all categories of birds_struck greater 0. So most of the collisions happen near ground level, which we have already seen at the beginning of chapter 3.1.1 that the flight phases near ground level (Take-Off, Climb, Approach and Landing Roll) have the highest number of strikes (see 3.1 Number of collisions by flight phase). For *birds_struck* equal to 0 the distribution is a little bit different from the others with a density of about 0.0016 from 200 to 500 foot. Remarkable is also the **peek for "Over 100" birds struck at a *height* of about 1800 foot**. For the "Over 100" category we have the lowest amount of observations in the dataset but nevertheless 4 out of 8 strikes happened here. So for this category we have the same probability of hitting birds near ground level and at about 1800 foot. One of the involved species here was a **flock of about 500 gulls**. Unfortunately the other observations had missing values in the variable "species", so we can not get more information about the species involved.

@@

There originally were only two levels in the data available for birds seen and we added one level for the missing values. The two plots at the top show the proportion of birds seen by the *time_of_day* (left plot) and the cloudiness of the *sky* (right plot). At the bottom you can see two spineplots that show the proportion of *birds_seen* along the two continuous variables *height* and *speed*.

@@

birds_seen_plt + birds_seenNo_sky
```
```{r echo=FALSE, fig.asp=0.4, fig.width = 7, fig.show="hold", fig.align = 'center'}
par(mfrow=c(1,2))
spineplot(birds$height, birds$birds_seen, ylab="", yaxlabels = "", xlab = "", col = rev(brewer.pal(n=3,name="BuGn")))
title(main = "Birds seen by height", cex.main=0.8, font.main= 3, col.main = "gray50", line = 0.5)
mtext(text = "height in foot",
      side = 1,#side 1 = bottom
      line = 2)

spineplot(birds$speed, birds$birds_seen,ylab="",yaxlabels = "", xlab = "",
          col = rev(brewer.pal(n=3,name="BuGn")))
title(main = "Birds seen by speed", cex.main=0.8, font.main= 3, col.main = "gray50",line = c(0.5))
#title(xlab="speed", line=4, cex.lab=1.2, family="Calibri Light")
mtext(text = "speed in knots",
      side = 1,#side 1 = bottom
      line = 2)
```

*Time of Day*: Based on the assumption that missing values in the data for *birds_seen* stands for the fact that the pilot did not see wildlife/birds before a strike happened, the average proportion (unweighted) of birds seen before a strike happened is 1-(0.65+0.69+0.65+0.87)/4=0.285. So most often pilots do not see birds before a collision. The proportion of overall **birds seen at night is much lower** with just 0.13 compared to the other times (Dawn, Day and Dusk).

*Sky*: The proportion of **wildlife seen increases with cloudiness**. This is a **contradiction** to what we would have expected. We would have thought that with increasing cloudiness the amount of wildlife/birds seen decreases, due to the fact that the pilot can not see as good as when there are less or no clouds. So one reason for this could be that most of the pilots aren't within the area of clouds when the collision happens, but have recorded the cloudiness of the sky. Another reason could be that birds fly more often when its cloudy.

From the spineplots we can see again that it is less likely to see birds before a collision, but the higher the *speed* & *height*, its becoming even less likely to see birds before a collision. For heights greater 2000 foot and for speed greater 140 knots, the probability to see birds before a collision decreases.

@@

In the following we will have a closer look on a self-generated variable, which compares the amount of *birds_seen* with the amount of *birds_struck*. It will be true if the number of birds struck is smaller than birds seen and false if not. In the beginning (chapter 3.1.1) we have already seen, that the number of *birds_seen* does not always infer that the same amount of birds were struck. By using this variable, we want to have a look at what influences the proportion of striking less birds than seen.

The top plot shows the proportion of less strikes than seen by phase of flight. On the bottom left we can see a comparison of the proportion of less strikes by the aircraft mass and on the bottom right we compare it by different number of engines.

```{r echo=FALSE, warning=FALSE,fig.asp=0.3}
#Create new column
#calculate only when seen is not na
seen_vs_strike <- birds[birds$birds_seen!="None" & !is.na(birds$birds_struck),]
seen_vs_strike$birds_struck <- factor(seen_vs_strike$birds_struck)
seen_vs_strike$birds_seen <- factor(seen_vs_strike$birds_seen)
seen_vs_strike$less_strike_than_seen <- as.factor(as.integer(seen_vs_strike$birds_struck) < as.integer(seen_vs_strike$birds_seen)+2)

less_strike_sums <- rep(addmargins(table(seen_vs_strike$less_strike_than_seen, seen_vs_strike$phase_of_flt),1)[3,3:8],each=2)
# phase of flight by birdsstruck < birds seen
seen_vs_strike_phase <- ggplot(seen_vs_strike[!is.na(seen_vs_strike$phase_of_flt) & as.integer(seen_vs_strike$phase_of_flt)>2,]) +
  geom_bar(aes(x=phase_of_flt, fill=less_strike_than_seen), position = "fill") +
      geom_line(aes(x=as.numeric(phase_of_flt)-2, y=..count../less_strike_sums[seq(1,12,2)], 
                group=less_strike_than_seen,
                color=less_strike_than_seen),
            stat = "count") +
  scale_color_manual(values = c("NA","black"))+
  guides(color = FALSE)+
  scale_fill_manual(values = color_pal_TF) +
  geom_text(aes(x=phase_of_flt, fill=less_strike_than_seen, 
                label=ifelse(round(..count../less_strike_sums,2) >
                               0.1,round(..count../less_strike_sums,2) , "")),
            stat='count',
            size = 3,
            position=position_fill(vjust = 0.5))+
  theme(legend.position="right", axis.title.y = element_blank())+
  theme.title+
  labs(title="Variables influencing less strikes than seen",
       subtitle = "Proportion of less strike than seen by phase of flight",
       fill='less strike \nthan seen'
  )

#ac_mass by less-strike-than-seen (MOSAIC)
seen_vs_strike_ac_mass <- ggplot(seen_vs_strike[!is.na(seen_vs_strike$ac_mass),]) +
  geom_mosaic(aes(x=product(ac_mass_renamed), fill=less_strike_than_seen)) +
  scale_fill_manual(values = color_pal_TF)+
  theme.title +
  theme(legend.position="none", 
        axis.text.y=element_blank(), 
        axis.ticks.y = element_blank(), 
        axis.text.x = element_text(angle = 90))+
  labs(subtitle = "Prop. of less strike than seen by aircraft mass")+
  ylab("") + 
  xlab("ac_mass in tons")

#num_engs by less-strike-than-seen (MOSAIC)
seen_vs_strike_num_engs <- ggplot(seen_vs_strike[!is.na(seen_vs_strike$num_engs),]) +
  geom_mosaic(aes(x=product(num_engs), fill=less_strike_than_seen)) +
  scale_fill_manual(values = color_pal_TF)+
  theme.title +
  theme(legend.position="none", 
        axis.text.y=element_blank(),
        axis.ticks.y = element_blank())+
  labs(subtitle = "Prop. of less strike than seen by number of engines")+
  ylab("")
